Tulai

### 2023-09-20 paper/paper_tulai.qmd
edit to simplify code up to the point where we read in the google sheet

```{r}
# load libraries
library(tidyverse)
library(here)
library(ggpmisc)
library(ggrepel)
library(ggbeeswarm)
```

@@

tl1 %>% 
  group_by(midpoint) %>% 
  tally() %>% 
  drop_na(midpoint) %>% 
ggplot() + 
  aes(midpoint, n) +
  geom_col()
```

@@

tl1 %>% 
  group_by(layer) %>% 
  tally() %>% 
  drop_na(layer) %>% 
ggplot() + 
  aes(layer, n) +
  geom_col()
```

tl2 <- tl1 %>%
  mutate(
    Blank = case_when(
      Breakage %in% c("0", "1", NA_character_) ~ Blank,

      TRUE ~ Breakage  
    )
  )

@@

rev(sort(table(tl2$Blank)))

# BM: still many typos and unusual categories in here, why is that? Can they be fixed?

@@

na_rows <- which(is.na(tl2$Blank))
print(na_rows)

zero_rows <- which(tl2$Blank == "0")
print(zero_rows)

one_rows <- which(tl2$Blank == "1")
print(one_rows)

rev(sort(table(tl3$Blank2))) # BM: still a lot of typos on the categories here, can you fix them?

@@

```{r}
# Remove rows based on row names
# tl3 <- tl3[!(rownames(tl3) %in% c("1030", "644", "543")), ]

# BM: why are we removing these rows? A comment here would be good to explain. A simpler way, by referencing the row numbers directly:
# SS: they are natural rocks, and don't want to count them as lithic

tl3 <- tl3[-c(1030, 644, 543), ]
```

@@

tl4 <- tl4 %>%
  # BM: the problem with this combination of mutate_all and ifelse that you use below is that they 
  # coerce all columns to be the same data type, so you'll see that all the numeric
  # columns are now character columns after you run these lines. So you need to
  # coerce those number columns back to numeric type so you can plot and compute
  # with them
  mutate_all(~ ifelse(is.na(.x) | .x == "", NA, .x)) %>% 
  # BM: what does this next line do?
  #SS: I wanted to remove any asterisks, whitespace, or *, ?, or !. is that wrong?

  mutate_all(~ str_replace(., "(?<=\\D)\\*(?=\\D)|\\s+|[*?!]+", "")) %>%
  # BM: what is this SHINE variable? I'm not familiar with it

  ##SS: we meant sickle shine, this is one of the criteria based on which Hole claimed the site is not related to agriculture (similar to ALi Kosh located in Dehluran)
  mutate(
    SHINE = str_trim(SHINE) %>% 
            str_to_title(),
    shine2 = SHINE
  ) %>%
  mutate(
    shine2 = case_when(
      SHINE %in% c("Subparallel", 
                   "Parallel", 
                   "Sub-Parallel", 
                   "Parallel-Subprarallel",
                   "Scaled", 
                   "Sub-Paralel", 
                   "Sub-Parallell", 
                   "Sub-Paallel",
                   "Sub-Paralell", 
                   "Sub-Prallel", 
                   "L:semiparallel/R:scaled",
                   "Semiparallel", 
                   "Semi-Parallel", 
                   "Scaled/Sub-Parallel",
                   "Seb-Parallel", 
                   "Semi-Abrupt") ~ NA_character_,
      TRUE ~ SHINE
    ),
    Utilization = if_else(Utilization == "0", NA_character_, Utilization),
    Retouch =     if_else(Retouch == "0", NA_character_, Retouch),
    SHINE =       if_else(SHINE == "0", NA_character_, SHINE),
    shine2 =      if_else(shine2 == "0", NA_character_, shine2),
    Blank2 =      if_else(Blank2 == "0", NA_character_, Blank2)
  ) %>% 
  mutate(
    SHINE = na_if(SHINE, "")
  )

# BM: I think this is a better way to inspect:
rev(sort(table(tl4$SHINE)))
rev(sort(table(tl4$shine2)))

@@

# Convert 'NA' and 'na' strings to actual NA
tl4$`Core Typology` <- na_if(tl4$`Core Typology`, "NA")
tl4$`Core Typology` <- na_if(tl4$`Core Typology`, "na")

@@

# Convert 'NA' and 'na' in new columns to actual NA
tl4$`core-typology` <- na_if(tl4$`core-typology`, "na")
tl4$`core-technology` <- na_if(tl4$`core-technology`, "na")

# Check the unique values again
# BM: still many typos, are they important?
rev(sort(table((tl4$`core-typology`))))
rev(sort(table((tl4$`core-technology`))))

@@

# BM: still some typos in there
rev(sort(table((tl4$`tool-typology`))))
rev(sort(table((tl4$`core-fragment`))))
```

```{r}
# Exploring data

tl4 %>% 
  # BM: we need this because of the mutate_all ifelse combo you have above
  mutate(midpoint = parse_number(midpoint)) %>% 
  group_by(midpoint) %>% 
  tally() %>% 
  drop_na(midpoint) %>% 
ggplot() + 
  aes(midpoint, n) +
  geom_col()

@@

tl_final <- tl4 %>%
  rename(
    blank =                Blank2,
   `sickle shine` =       shine2,
    length =              `Length (mm)` ,
    width =               `Width (mm)`,
    thickness =           `Thickness (mm)`, 
    typology =             Typology,
    utilization =          Utilization,
    retouch =              Retouch,
    blank =                Blank2, 
   `blank part` =         `Blank_part`,
   `bulb thickness` =     `Bulb Thickness (mm)`, 
   `platform thickness` = `Platform Thickness (mm)`, 
   `platform length` =    `platform Length (mm)`, 
    eraillure =            Eraillure , 
    cortex =               `Cortex (%)`)  %>%
  mutate(al = paste0(toupper(area), "_", layer)) %>%
  relocate(al, .after = layer)   %>%
  select(lithic_id, 
         TA, 
         area, 
         layer,
         al,
         cortex,
         depth, upper, lower, midpoint, 
        `length`, `width`, `thickness`, 
         typology, 
         utilization, 
        `sickle shine`,
         retouch, 
        `typology-tool-core`,
        `tool-typology`,
         blank, `blank part`, 
        `core-typology`, `core-fragment`,`core-technology`, 
        cortex)  %>% 
# Update the 'typology-tool-core' column based on 'Retouch'
  mutate(
    `typology-tool-core` = case_when(
      retouch == 1 ~ "tool",           
      utilization == 1 ~ "tool",
      `sickle shine` == 1 ~ "tool",

      TRUE ~ `typology-tool-core`      
    )
  ) %>% 
  mutate(
    `tool-typology` = case_when(
      utilization == 1 ~ "utilized tool",
      `sickle shine` == 1 ~ "sickle shine",
      TRUE ~ `tool-typology`               
    )
  ) %>% 
  mutate(
    `tool-typology` = case_when(
     # If Retouch is 1 and tool-typology is NA
       retouch == 1 & is.na(`tool-typology`) ~ "retouched piece",
       `sickle shine` == 1 & is.na(`tool-typology`) ~ "sickle shine",

@@

# BM: take a look at tool typology
rev(sort(table(tl_final$`tool-typology`)))

@@

# Convert each column to numeric
for (col_name in columns_to_convert) {
  # Replace non-numeric characters with NA
  # BM: this is quite an unusual method for doing this, 
  # where did you find it?
  tl_final[[col_name]][tl_final[[col_name]] %in% c("", "NA", "N/A")] <- NA
  # Convert to numeric
  tl_final[[col_name]] <- as.numeric(tl_final[[col_name]])
}

# BM: this is how I would convert those columns to numeric, 
# I think it's easier to read and requires less typing
# tl_final %>% 
#  mutate(across(all_of(columns_to_convert), parse_number))

# check the structure to see if they have been converted
str(tl_final)
```

@@

tl_final <- tl_final %>%
  rowwise() %>%
  mutate(utilization = ifelse(is.na(`sickle shine`) | is.na(utilization), utilization,
                              ifelse(`sickle shine` == 1 & utilization == 1, NA_real_, utilization))) %>%
  ungroup()

@@

```{r}
#number of `sickle shine` in clumn `sickle shine`=1 and tooltypegroup are not match: 
rev(sort(table(tl_final$tooltypegroup)))
rev(sort(table(tl_final$`sickle shine`)))

#troubleshoot:
tl_final %>% filter(`sickle shine` == 1) %>% 
  select(`tool-typology`, 
         `sickle shine`)

tl_final %>% filter(`tool-typology` == "sickle shine") %>% 
  select(`tool-typology`, `sickle shine`)

tl_final <- tl_final %>%
  mutate(
    `tool-typology` = case_when(
      `sickle shine` == 1 ~ "sickle shine",
      utilization == 1 ~ "utilized tool",
      TRUE ~ `tool-typology`               
    )
  )

rev(sort(table(tl_final$`tool-typology`)))
```

@@

# Checking the unique values of tooltypegroup
rev(sort(table(tl_final$tooltypegroup)))

@@

library(dplyr)
library(stringr)

@@

# View the first few rows to confirm
head(tl_final)
rev(sort(table(tl_final$tooltypegroup)))
rev(sort(table(tl_final$retouch)))
```

@@

# take a look
rev(sort(table(tl_final$core)))
rev(sort(table(tl_final$corefrag)))
rev(sort(table(tl_final$`core-fragment`)))
rev(sort(table(tl_final$blank)))
```

Summary tables: 1. AREA

@@

# BM: looks like this is area and layer, is that right? I see c1, d1, d2, d3, d4 etc.
# SS: those are different excavated areas, we mainly consider ta, tp1, and d1.
knitr::kable(summary_table_area)

```{r}
library(kableExtra)

kable(summary_table_area, 
      caption = "Table: Summary by Area", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)

```{r}
library(dplyr)

@@

library(kableExtra)

kable(summary_table_area_filtered, 
      caption = "Table: Summary by Area for 'ta', 'd1', and 'tp1'", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)

@@

2.LAYER

@@

```{r}
# Generate the table
kable(summary_table_layer, 
      caption = "Table: Summary by Layer", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)

@@

AREA-LAYER:

@@

kable(summary_table_al, 
      caption = "Table: Summary by Al", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)

@@

Techno-Typo:

@@

rev(sort(table(tl_final$tooltypegroup)))
```

@@

summary_table_tooltype <- tl_final %>%
  filter(tooltypegroup != 'utilized tool') %>%  # Exclude 'utilized tool'
  group_by(tooltypegroup) %>% 
  summarise(

```{r}
#generate a table

@@

kable(
  summary_table_al,
  col.names = c(
    "Area-Layer",
    "Blade",
    "Flake",
    "Bladelet",
    "Tools",
    "Cores",
    "Core Rejuvenations",
    "Retouched Tools",
    "Utilized Tools",
    "Sickle Shine",
    "Total"
  ),
  caption = "Table: Summary ",
  align = 'c'
) %>%
  kable_styling(
    bootstrap_options = c("striped", 
                          "hover", 
                          "condensed"),
    full_width = FALSE,
    font_size = 12
  ) %>%
  row_spec(0, 
           bold = TRUE, 
           font_size = 14)

@@

kable(
  summary_table_tool_area,
  col.names = c("Tools", "D1", "TA", "TP1", "Total"),
  caption = "Table: Tool Typology in three different main areas.",
  align = 'c'
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F,
    font_size = 12
  ) %>%
  column_spec(1, extra_css = "text-align:left;") %>%
  row_spec(0, bold = TRUE, font_size = 14)

@@

# Generate the table
library(kableExtra)

kable(summary_table_depth_area_layer, 
      caption = "Table: Summary by Depth, Area, and Layer", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)
```

@@

# Generate the table
library(kableExtra)

kable(summary_table_depth_area_layer, 
      caption = "Table: Summary by Depth, Area, and Layer", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)

@@

# Generate the table
library(kableExtra)

kable(summary_table_depth_area_layer, 
      caption = "Table: Summary by Depth, Area, and Layer", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)

@@

#Here, I've imported the table 'Summary by Depth, Area, Layer' into Google Sheets for easier management.
#I thought this format would simplify the process of those depths needed to be splitting up.

>>>>

```{r}
# load libraries
library(tidyverse)
library(here)
library(ggpmisc)
library(ggrepel)
library(ggbeeswarm)
library(cowplot)
```

@@

p1 <- 
tl1 %>% 
  group_by(midpoint) %>% 
  tally() %>% 
  drop_na(midpoint) %>% 
ggplot() + 
  aes(midpoint, n) +
  geom_col()

p2 <- 
tl1 %>% 
  group_by(layer) %>% 
  tally() %>% 
  drop_na(layer) %>% 
ggplot() + 
  aes(layer, n) +
  geom_col()

plot_grid(p1, p2, ncol = 1)
```

@@

tl2 <- tl1 %>%
  mutate(
    Blank = case_when(
      Breakage %in% c("0", "1", NA_character_) ~ Blank, 
      TRUE ~ Breakage  
    )
  )
```

@@

# rev(sort(table(tl2$Blank)))

# BM: still many typos and unusual categories in here, why is that? Can they be fixed? This might increase our sample size.

@@

na_rows <- which(is.na(tl2$Blank))

zero_rows <- which(tl2$Blank == "0")
one_rows <- which(tl2$Blank == "1")

# rev(sort(table(tl3$Blank2)))

@@

```{r}
# SS: they are natural rocks, and don't want to count them as lithic
# BM: better to remove them by their ID than by row number, since row number can easily change, but ID cannot

tl3 <- tl3 %>% 
  filter(!lithic_id %in% c("3154", "1741", "1680"))
```

@@

tl4 <- tl4 %>%
  # BM: I updated this to use across
  mutate(across(everything(), 
                ~ifelse(is.na(.x)|.x == ""|.x == "-", NA, .x))) %>% 
  #SS:  remove any asterisks, whitespace, or *, ?, or !. is that wrong?
   mutate(across(everything(),
                 ~str_replace(., "(?<=\\D)\\*(?=\\D)|\\s+|[*?!]+", ""))) %>%
  ##SS: we meant sickle shine, this is one of the criteria based on which Hole claimed the site is not related to agriculture (similar to ALi Kosh located in Dehluran)
  mutate(
    SHINE = str_trim(SHINE) %>% 
            str_to_title(),
    shine2 = SHINE
  ) %>%
  mutate(
    shine2 = case_when(
      SHINE %in% c("Subparallel", 
                   "Parallel", 
                   "Sub-Parallel", 
                   "Parallel-Subprarallel",
                   "Scaled", 
                   "Sub-Paralel", 
                   "Sub-Parallell", 
                   "Sub-Paallel",
                   "Sub-Paralell", 
                   "Sub-Prallel", 
                   "L:semiparallel/R:scaled",
                   "L:semiparallel/R: Scaled",
                   "Semiparallel", 
                   "Semi-Parallel", 
                   "Scaled/Sub-Parallel",
                   "Seb-Parallel", 
                   "Semi-Abrupt") ~ NA_character_,
      TRUE ~ SHINE
    ),
    Utilization = if_else(Utilization == "0", NA_character_, Utilization),
    Retouch =     if_else(Retouch == "0", NA_character_, Retouch),
    SHINE =       if_else(SHINE == "0", NA_character_, SHINE),
    shine2 =      if_else(shine2 == "0", NA_character_, shine2),
    Blank2 =      if_else(Blank2 == "0", NA_character_, Blank2)
  ) %>% 
  mutate(
    SHINE = na_if(SHINE, "")
  )

# BM: I think this is a better way to inspect:
# rev(sort(table(tl4$SHINE)))
# rev(sort(table(tl4$shine2)))

@@

@@

# Check the unique values again
# rev(sort(table((tl4$`core-typology`))))
# rev(sort(table((tl4$`core-technology`))))

@@

# BM: still some typos in there
# rev(sort(table((tl4$`tool-typology`))))
# rev(sort(table((tl4$`core-fragment`))))
```

@@

tl_final <- tl4 %>%
  rename(
    blank =                Blank2,
   `sickle shine` =       shine2,
    length =              `Length (mm)` ,
    width =               `Width (mm)`,
    thickness =           `Thickness (mm)`, 
    typology =             Typology,
    utilization =          Utilization,
    retouch =              Retouch,
    blank =                Blank2, 
   `blank part` =         `Blank_part`,
   `bulb thickness` =     `Bulb Thickness (mm)`, 
   `platform thickness` = `Platform Thickness (mm)`, 
   `platform length` =    `platform Length (mm)`, 
    eraillure =            Eraillure , 
    cortex =               `Cortex (%)`)  %>%
  mutate(al = paste0(toupper(area), "_", layer)) %>%
  relocate(al, .after = layer)   %>%
  select(lithic_id, 
         TA, 
         area, 
         layer,
         al,
         cortex,
         depth, upper, lower, midpoint, 
        `length`, `width`, `thickness`, 
         typology, 
         utilization, 
        `sickle shine`,
         retouch, 
        `typology-tool-core`,
        `tool-typology`,
         blank, `blank part`, 
        `core-typology`, `core-fragment`,`core-technology`, 
        cortex)  %>% 
# Update the 'typology-tool-core' column based on 'Retouch'
  mutate(
    `typology-tool-core` = case_when(
      retouch == 1 ~ "tool",           
      utilization == 1 ~ "tool",
      `sickle shine` == 1 ~ "tool",
      TRUE ~ `typology-tool-core`      
    )
  ) %>% 
  mutate(
    `tool-typology` = case_when(
      utilization == 1 ~ "utilized tool",
      `sickle shine` == 1 ~ "sickle shine",
      TRUE ~ `tool-typology`               
    )
  ) %>% 
  mutate(
    `tool-typology` = case_when(
     # If Retouch is 1 and tool-typology is NA
       retouch == 1 & is.na(`tool-typology`) ~ "retouched piece",
       `sickle shine` == 1 & is.na(`tool-typology`) ~ "sickle shine",

@@

# BM: take a look at tool typology
# rev(sort(table(tl_final$`tool-typology`)))

@@

# Convert each column to numeric

# BM: this is how I would convert those columns to numeric, 
# I think it's easier to read and requires less typing
tl_final <- 
tl_final %>% 
 mutate(across(all_of(columns_to_convert), 
               parse_number))

# check the structure to see if they have been converted
# str(tl_final)
```

@@

tl_final <- tl_final %>%
  rowwise() %>%
  mutate(utilization = ifelse(is.na(`sickle shine`) | is.na(utilization),
                              utilization,
                              ifelse(`sickle shine` == 1 & utilization == 1, 
                                     NA_real_, 
                                     utilization))) %>%
  ungroup() %>%
  # make the number of `sickle shine` in column `sickle shine` match:
  mutate(
    `tool-typology` = case_when(
      `sickle shine` == 1 ~ "sickle shine",
      utilization == 1 ~ "utilized tool",
      TRUE ~ `tool-typology`               
    )
  )

@@

@@

# Checking the unique values of tooltypegroup
# rev(sort(table(tl_final$tooltypegroup)))

@@

@@

# View the first few rows to confirm
# head(tl_final)
# rev(sort(table(tl_final$tooltypegroup)))
# rev(sort(table(tl_final$retouch)))
```

@@

# take a look
# rev(sort(table(tl_final$core)))
# rev(sort(table(tl_final$corefrag)))
# rev(sort(table(tl_final$`core-fragment`)))
# rev(sort(table(tl_final$blank)))
```

### Summary tables: AREA

@@

# SS: we mainly consider ta, tp1, and d1.
knitr::kable(summary_table_area,
             caption = "Summary of lithics in all areas")
```

@@

knitr::kable(summary_table_area_filtered,
             caption = "Summary of lithics in the main areas")

@@

### Summary tables: LAYER

@@

```{r}
# Generate the table
kable(summary_table_layer, 
      caption = "Summary of lithics by Layer", 
      align = 'c')

@@

### Summary tables: AREA-LAYER:

@@

kable(summary_table_al, 
      caption = "Summary of lithics by Area and layer")

@@

### Summary tables: Techno-Typology

@@

# rev(sort(table(tl_final$tooltypegroup)))
```

@@

summary_table_tooltype <- tl_final %>%
  filter(tooltypegroup != 'utilized tool') %>%  # Exclude 'utilized tool'
  group_by(tooltypegroup) %>% 
  summarise(
  )
```

```{r}
# generate a table

@@

kable(
  summary_table_al,
  col.names = c(
    "Area-Layer",
    "Blade",
    "Flake",
    "Bladelet",
    "Tools",
    "Cores",
    "Core Rejuvenations",
    "Retouched Tools",
    "Utilized Tools",
    "Sickle Shine",
    "Total"
  ),
  caption = "Summary of lithics by techno-typological classification")

@@

kable(
  summary_table_tool_area,
  col.names = c("Tools", "D1", "TA", "TP1", "Total"),
  caption = "Tool Typology in the three main areas")

@@

# Generate the table
kable(summary_table_depth_area_layer, 
      caption = "Summary of lithics by Depth, Area, and Layer")
```

@@

# Generate the table

kable(summary_table_depth_area_layer, 
      caption = "Summary of lithics by Depth, Area, and Layer")

@@

# Generate the table
kable(summary_table_depth_area_layer, 
      caption = "Summary of lithics by Depth, Area, and Layer")

@@

# import table with depths that we have subdivided
depths_divided_tbl <- read_csv(here("data/Lookup Table - Sheet1.csv"))

# join with lithic data
tl_final %>% 
  left_join(depths_divided_tbl) %>% View

```{r}
#| eval: false

# Here, I've imported the table 'Summary by Depth, Area, Layer' into Google Sheets for easier management.
#I thought this format would simplify the process of those depths needed to be splitting up.
